Manuscript for foodborne-outbreak Project

### 2023-03-16 code/analysis_code/exploratory_analysis.qmd
update

Norovirus_Hosp=training_data_final%>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
  filter(Simplified_Etiology=="Salmonella")%>%
  filter(`IFSAC Category`!= %n% C("Multiple","Other"))%>%
  count()%>%
  mutate(total_hospitalizations=n*Hospitalizations)

>>>>

Norovirus_Hosp=training_data_final%>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
  filter(Simplified_Etiology=="Salmonella")%>%
  filter(`IFSAC Category` != "Multiple")%>%
  filter(`IFSAC Category` != "Other") %>%
  count()%>%
  mutate(total_hospitalizations=n*Hospitalizations)

### 2023-03-16 code/analysis_code/statistical_analysis.qmd
update

# Load libraries
```{r}
library(here)
library(tidyverse)
library(tidymodels)
library(poissonreg)
```
# Load data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
glimpse(training_data_final)
```
## First health outcome of interest: Illnesses
### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
```{r}
# Checking the distribution of Illnesses
training_data_final%>%
  ggplot(aes(Illnesses))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
image_path <- here("results")
ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")
# Creating a tibble only includes our predictors of interest
illness_data=training_data_final%>%
  select(c(1,4,5,7,10,13,14))%>%
  mutate_if(is.character, as.factor)
glimpse(illness_data)
```
### Modeling `Illness` using linear regression and Poisson regression
#### Since the outcome is counts and the event ocurrs independently, it follows a Poisson distribution
```{r}
# Model specification and fitting for linear regression model
lm_spec=linear_reg()%>%
  set_engine(engine="lm")
# Modeling with main predictor `Simplified_Etiology`
lm_fit1=lm_spec%>%
  fit(Illnesses~Simplified_Etiology, data=illness_data)
# Modeling with all predictors
lm_fit2=lm_spec%>%
  fit(Illnesses~., data=illness_data)
# Model specification and fitting for Poisson regression model
# create a workflow for poisson regression model, we will use the default glm engine
# Modeling with main predictor `Simplified_Etiology`
poisson_wf1=workflow(Illnesses~Simplified_Etiology, poisson_reg())
# Modeling with all predictors
poisson_wf2=workflow(Illnesses~., poisson_reg())
# Fitting models
poisson_fit1=fit(poisson_wf1,data=illness_data)
poisson_fit2=fit(poisson_wf2,data=illness_data)
```

### Model Evaluation
```{r}
# Checking parameters for linear regression model
tidy(lm_fit1) %>%
  arrange(-estimate)
tidy(lm_fit2)%>%
  arrange(-estimate)
# Checking parameters for Poisson regression model
tidy(poisson_fit1)%>%
 arrange(-estimate)
tidy(poisson_fit2)%>%
 arrange(-estimate)
# Model evaluation
Result=lm_fit1%>%
 predict(new_data=illness_data)%>%
 mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_single")%>%
bind_rows(lm_fit1%>%
            predict(new_data=illness_data)%>%
            mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_multiple")%>%
            bind_rows(poisson_fit1%>%
                        predict(new_data= illness_data)%>% 
                        mutate (true_value=illness_data$Illnesses, model= "glm_single")%>%
                        bind_rows(poisson_fit2%>%
                                    predict(new_data= illness_data)%>% 
                                    mutate (true_value=illness_data$Illnesses, model= "glm_multiple")
                        )
            )
)
Result%>%
  group_by(model)%>%
  rmse(truth=true_value,estimate=.pred)

```
### According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
## Second health outcome of interest: Hospitilization
### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
```{r}
# Checking the distribution of Hospitalization
training_data_final%>%
  ggplot(aes(Hospitalizations))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
# Creating a tibble only includes our predictors of interest
hosp_data=training_data_final%>%
  select(c(1,4,5,7,11,13,14))%>%
  mutate_if(is.character, as.factor)
glimpse(hosp_data)
ggsave("histogram_Hospitalizations.png",path=image_path)
```
### Modeling `Hospitalizations` using linear regression and Poisson regression
```{r}
# Model specification and fitting for linear regression model
lm_spec=linear_reg()%>%
  set_engine(engine="lm")
# Modeling with main predictor `Simplified_Etiology`
lm_fit3=lm_spec%>%
  fit(Hospitalizations~Simplified_Etiology, data=hosp_data)
# Modeling with all predictors
lm_fit4=lm_spec%>%
  fit(Hospitalizations~., data=hosp_data)
# Model specification and fitting for Poisson regression model
# create a workflow for poisson regression model, we will use the default glm engine
# Modeling with main predictor `Simplified_Etiology`
poisson_wf3=workflow(Hospitalizations~Simplified_Etiology, poisson_reg())
# Modeling with all predictors
poisson_wf4=workflow(Hospitalizations~., poisson_reg())
# Fitting models
poisson_fit3=fit(poisson_wf3,data=hosp_data)
poisson_fit4=fit(poisson_wf4,data=hosp_data)
```

### Model evaluation
```{r}
Result=lm_fit3%>%
 predict(new_data=hosp_data)%>%
 mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_single")%>%
bind_rows(lm_fit4%>%
            predict(new_data=hosp_data)%>%
            mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_multiple")%>%
            bind_rows(poisson_fit1%>%
                        predict(new_data= hosp_data)%>% 
                        mutate (true_value=hosp_data$Hospitalizations, model= "glm_single")%>%
                        bind_rows(poisson_fit2%>%
                                    predict(new_data= hosp_data)%>% 
                                    mutate (true_value=hosp_data$Hospitalizations, model= "glm_multiple")
                        )
            )
)
Result%>%
  group_by(model)%>%
  rmse(truth=true_value,estimate=.pred)

@@

>>>>

# Load libraries
```{r}
# Pathing
library(here)

# Data Handling and Modeling
library(tidyverse)
library(tidymodels)
library(poissonreg)
```
# Load Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
glimpse(training_data_final)
```
# First Health Outcome of Interest: Illnesses

Predictors
-   Time (Year)
-   Pathogen Type
-   Etiology
-   Location
-   State
-   Food Type (IFSAC Category)

```{r}
# Checking Illness Distribution
training_data_final %>%
  ggplot(aes(Illnesses))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
image_path <- here("results")
ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")

# Tibble with Predictors of Interest
illness_data=training_data_final %>%
  select(c(1, 4, 5, 7, 10, 13, 14)) %>%
  mutate_if(is.character, as.factor)
glimpse(illness_data)
```

## Modeling Illness with Linear Regression and Poisson Regression

Since the outcome is counts/frequency and the event occurs independently, it follows a Poisson distribution. To create a workflow for poisson regression model, the default *glm* engine will be used.

```{r}
# Linear Regression
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")
## Modeling with Main Predictor `Simplified_Etiology`
lm_fit1 = lm_spec %>%
  fit(Illnesses ~ Simplified_Etiology, data = illness_data)
## Modeling with All Predictors
lm_fit2 = lm_spec %>%
  fit(Illnesses ~ ., data = illness_data)

# Poisson Regression
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf1 = workflow(Illnesses ~ Simplified_Etiology, poisson_reg())
## Modeling with All Predictors
poisson_wf2 = workflow(Illnesses ~ ., poisson_reg())
## Model Fitting
poisson_fit1 = fit(poisson_wf1, data=illness_data)
poisson_fit2 = fit(poisson_wf2, data=illness_data)
```

### Model Evaluation
```{r}
# Checking Parameters for Linear Regression Model
tidy(lm_fit1) %>%
  arrange(-estimate)
tidy(lm_fit2) %>%
  arrange(-estimate)

# Checking Parameters for Poisson Regression Model
tidy(poisson_fit1) %>%
 arrange(-estimate)
tidy(poisson_fit2) %>%
 arrange(-estimate)

# Model Evaluation
Result = lm_fit1 %>% 
  predict(new_data = illness_data) %>% 
  mutate(.pred = round(.pred), 
        true_value = illness_data$Illnesses, model= "lm_single") %>% 
  bind_rows(lm_fit1 %>% 
              predict(new_data=illness_data) %>% 
              mutate(.pred=round(.pred), 
                     true_value = illness_data$Illnesses, model= "lm_multiple") %>% 
              bind_rows(poisson_fit1 %>% predict(new_data= illness_data) %>%
                          mutate(true_value=illness_data$Illnesses, 
                                 model= "glm_single") %>% 
                          bind_rows(poisson_fit2 %>% 
                                      predict(new_data= illness_data)%>% 
                                      mutate(true_value=illness_data$Illnesses, 
                                             model= "glm_multiple")))
)

Result <- Result %>%
  group_by(model) %>%
  rmse(truth=true_value, estimate = .pred) %>%
  rename(illness_model = model)

saveRDS(Result, file = here("results", "illness_model_evaluation.rds"))

```
According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.

# Second Health Outcome of Interest: Hospitilization

Predictors
-   Time (Year)
-   Pathogen Type
-   Etiology
-   Location
-   State
-   Food Type (IFSAC Category)

```{r}
# Checking the Distribution of Hospitalization
training_data_final %>%
  ggplot(aes(Hospitalizations)) +
  geom_histogram() +
  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
# Tibble for Variables of Interest
hosp_data = training_data_final %>%
  select(c(1, 4, 5, 7, 11, 13, 14))%>%
  mutate_if(is.character, as.factor)
glimpse(hosp_data)
ggsave("histogram_Hospitalizations.png", path = image_path)
```

## Modeling `Hospitalizations` using Linear Regression and Poisson Regression

To create a workflow for the Poisson regression model, we will use the default *glm* engine again.

```{r}
# Model Specification and Fitting for Linear Regression Model
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")
## Modeling with Main Predictor `Simplified_Etiology`
lm_fit3 = lm_spec %>%
  fit(Hospitalizations ~ Simplified_Etiology, data = hosp_data)
## Modeling with All Predictors
lm_fit4 = lm_spec %>%
  fit(Hospitalizations ~ ., data = hosp_data)

# Model Specification and Fitting for Poisson Regression Model
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf3 = workflow(Hospitalizations ~ Simplified_Etiology, poisson_reg())
## Modeling with all predictors
poisson_wf4=workflow(Hospitalizations~., poisson_reg())
## Fitting models
poisson_fit3 = fit(poisson_wf3,data = hosp_data)
poisson_fit4 = fit(poisson_wf4,data = hosp_data)
```

### Model Evaluation
```{r}
Result = lm_fit3 %>%
 predict(new_data = hosp_data) %>%
 mutate(.pred = round(.pred), true_value = hosp_data$Hospitalizations,
        model= "lm_single") %>% 
  bind_rows(lm_fit4 %>% 
              predict(new_data=hosp_data) %>% 
              mutate (.pred=round(.pred), 
                      true_value=hosp_data$Hospitalizations, 
                      model= "lm_multiple") %>% 
              bind_rows(poisson_fit1 %>% 
                        predict(new_data = hosp_data) %>%
                        mutate(true_value = hosp_data$Hospitalizations, 
                               model= "glm_single") %>% 
                          bind_rows(poisson_fit2 %>% 
                                      predict(new_data= hosp_data) %>% 
                                      mutate(true_value = hosp_data$Hospitalizations, 
                                             model= "glm_multiple")))
)

Result <- Result %>%
  group_by(model)%>%
  rmse(truth = true_value, estimate = .pred) %>%
  rename(hospital_model = model)

saveRDS(Result, file = here("results", "hospital_model_evaluation.rds"))

@@

According to the result, the model built using linear regression with multiple predictors seems to have the smallest root mean squared error.

### 2023-03-16 products/manuscript/Manuscript.qmd
update

We will be building regression models using these predictors and health outcomes as well; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.

@@

## Basic Statistical Analysis

>>>>

Afterwards, we built regression models using these predictors and health outcomes; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.

@@

## Basic Statistical Analysis
Multivariate Poisson regression performed for illness as the health outcome proved to be the best model (RMSE = 55.6) among simple linear, multiple linear, simple Poisson, and multivariate Poisson regressions.

```{r echo=FALSE}
illness_model_evaluated <- readRDS(here("results", "illness_model_evaluation.rds"))
rmarkdown::paged_table(illness_model_evaluated)
```

For hospitalization as the health outcome, multivariate linear regression was the best model (RMSE = 7.09) among those previously listed.

```{r echo=FALSE}
illness_model_evaluated <- readRDS(here("results", "hospital_model_evaluation.rds"))
rmarkdown::paged_table(illness_model_evaluated)
```
